---
title: "Detailed Analysis of High ECV Cohort"
author: "E Bryan Crenshaw III"
date: "April 7, 2015"
output: html_document
---

**Objective**: Subset the patients with high Ear Canal Volume, and examine their record more fully in [AudGenDB 2.0](https://audgendb.chop.edu/app2/).  

**Approach**: Load 'tympnorms.sanstubes_mid' dataset of patients with PTA between 0 and 15 dB from [Tympanometry Download](Tympanometry_Download.html).  Subset the data to identify those patients with [high ECVs according to CHOP standards](data/Tympanometry226HzNorms.pdf). Ouput the Patient.Alias column to a text file, and then take a look at their data in AudGenDB:


```{r LoadDataSubset}
# Data from patients with normal PTAs
load("data/tympnorms_sanstubes_mid.RData")
# Isolate data from patients with high ECV
tymp_hiecv <- subset(tympnorms.sanstubes_mid, volume > 1.5,)
write(unique(tymp_hiecv$Patient.Alias), file = "data/tymp_ol.txt")
print(tymp_hiecv$Patient.Alias)
```

The text from 'tymp_ol.txt' was input into AudGenDB, and the data was quickly [analyzed in Excel](data/TympECVgt1-5_all-2015-04-07 16-00-51.515449-data.xlsx).  The data were sorted by PTA from largest to smallest.  All of those rows of data with PTA higher than 15 dB were highlighted in yellow, and the rows were resorted by Patient.Alias.  Highlighted rows were dispersed broadly in the dataset.

```{r QuickStats}
tymp_hiecv.patientno <- length(unique(tymp_hiecv$Patient.Alias))
```

###Observations###

Of the `r tymp_hiecv.patientno` patients with high ECV, many of the high ECV patients had progressive or variable hearing loss.  

**Next Steps**: Download the data again, including audiological parameters. Subsequently, identify patients that have ever had audiological anomalies despite some tests showing normal PTAs (start with PTA > 15 dB). Eliminate these patients from the dataset, and rerun analyses.  


<br /><br /><br />
