---
title: "Static Compliance Probability Density Function By Age"
author: "E Bryan Crenshaw III with Cody Bender"
date: "March 31, 2017"
output: html_document

---
**Objective**: Plot the probability density function of static compliance in three age groups (young = age >= 0.5 & age < 3; middle = age >= 3 & age <= 15; old = (age >= 15 & age <= 21 ). 

**Approach**: Loaded 'nodupall', a dataset of patients from AudGenDB with no 
tympnaostomy CPT codes checked against a list of tympanostomy patients. Removed 
duplicate entries with the same patient alias. Patients with missing values for ear canal volume, middle ear pressure, static compliance, or age were also removed. Patients were then divided into 3 age groups: young (0.65-3 years), middle (3-15 years), and older (15-21 years). Quantiles were calculated for each group and probability density functions were plotted.

```{r}
#Data divided into 3 age groups
load(file = "nodupall.Rdata")
nodupall_yng <- subset(nodupall, age >= 0.5 & age < 3,)
nodupall_mid <- subset(nodupall, age >= 3 & age < 15,)
nodupall_old <- subset(nodupall, age >= 15 & age < 21,)
```
```{r, echo=FALSE, warning=FALSE}
library(psych)
```
**Basic Statistical Analyses:**
```{r}
nodupalls_yng.describe <- describe(nodupall_yng$sc)
print("Basic Statistics on Static Compliance of Children 6 months to <3 years Old :"); nodupalls_yng.describe
nodupalls_mid.describe <- describe(nodupall_mid$sc)
print("Basic Statistics on Static Compliance of Children 6 months to <3 years Old :"); nodupalls_mid.describe
nodupalls_old.describe <- describe(nodupall_old$sc)
print("Basic Statistics on Static Compliance of Children 6 months to <3 years Old :"); nodupalls_old.describe
```

**Quantile Analyses:**
```{r}
# Show quantiles of percents in the probs vector below
sc_yng <- quantile(nodupall_yng$sc,  probs = c(2.5, 5, 10, 50, 90, 95, 97.5, NA)/100, na.rm = TRUE)
print("Quantiles of Younger Children:"); print(sc_yng);
sc_mid <- quantile(nodupall_mid$sc,  probs = c(2.5, 5, 10, 50, 90, 95, 97.5, NA)/100, na.rm = TRUE)
print("Quantiles of Middle Children:"); print(sc_mid)
sc_old <- quantile(nodupall_old$sc,  probs = c(2.5, 5, 10, 50, 90, 95, 97.5, NA)/100, na.rm = TRUE)
print("Quantiles of Older Children:"); print(sc_old)
```

```{r plot} 
# Plot the sc of the 3 age ranges with outliers removed
plot(density(nodupall_yng$sc, na.rm = TRUE), main = "Static Compliance of Children with PTA <=15 dB & No Pathology", xlab = "Static Compliance (mL)", xlim = c(0.0,1.70), ylim = c(0,2.5))
lines(density(nodupall_mid$sc, na.rm = TRUE), col =2)
lines(density(nodupall_old$sc, na.rm = TRUE), col = 3)
legend("topright", c("Young (0.5-<3yo)", "Middle (3-15yo)","Older (15-21yo)"), cex=1.1, col=c("black","red","green"), pch=21:23, lty=1:3)
```

































**Observations**: The dataset contained 1802 patients after additional tympanograms performed for each patient after the first were removed. The vast majority of patients fell into the Middle subset between ages 3-15.


**Comparisons By Age**:

The Younger subset contained 57 patients and had a 90% range of static compliance from 0.200 to 0.800.

The Middle subset contained 1556 patients and had a 90% range of static compliance from 0.200 to 1.300.

The Older subset contained 189 patients and had a 90% range of static compliance from 0.300 to 2.106.